CytoGAN: Generative Modeling of Cell Images

### 2017-12-01 paper/paper.tex
Small update to paper and poster
Fixed a few things in the paper

\section{Introduction}
Advances in high-throughput microscopy systems have enabled acquisition of large volumes of high-resolution cell images \cite{caicedo2017data}. This paves the way for novel computational methods that can leverage these large quantities of data to study biological systems. Our work focuses on the task of morphological profiling, which aims to map microscopy images of cells to representation vectors that capture salient features and axes of variation in an unsupervised manner \cite{caicedo_profiling}. These representations ideally divide the morphological space into clusters of cells with similar properties or, in the case of induced perturbations, similar function.

@@

We employ Generative Adversarial Networks (GANs) \cite{goodfellow2014generative} to build a generative model of single cell images from the BBBC021 dataset \cite{Ljosa2012}. We show that this model learns rich feature representations and synthesizes realistic images that are useful for exploring morphological variation in cells. The main advantages of this method are:
\begin{itemize}
  \item \emph{Adaptability}: Unsupervised representation learning, which can readily incorporate new data. This is important given that applications of morphological profiling usually lack ground truth annotations, but instead aim to reveal the structure of the image collection.
  \item \emph{Specialization}: Because this method learns from its training data it is able to disentangle factors of variation and extract inherent semantic relationships. Transfer learning approaches lack this ability and cannot capture the intrinsic relations between the biologically meaningful channels.
  \item \emph{Translating into biological phenotypes}: The generative abilities of this model enable useful visualizations of cells to help translate data variations into meaningful biological phenotypes.
\end{itemize}

@@

\section{Exploring Biological Phenotypes Using Cell Image Synthesis}\label{synthesis}
GANs have been proven to synthesize realistic images both within and beyond the biological domain \cite{goodfellow2014generative,radford2015unsupervised,osokin2017gans}. We examine if this ability transfers to cell images extracted from the BBBC021 dataset of human breast cancer cell lines. Each image in BBBC021 consists of three \emph{channels} corresponding to DNA, F-Actin and $\beta$-Tubulin, cellular components visualized by fluorescence microscopy. We stack these channels and treat them as RGB images via a simple $\text{DNA} \mapsto \text{R}, \text{$\beta$-Tubulin} \mapsto \text{G}, \text{F-Actin} \mapsto \text{B}$ mapping. We obtain ca. 1.3 million single cell images by segmenting the raw images using CellProfiler. Finally, we normalize the luminance of each channel to the range [0 -- 1].

Figure \ref{fig:generated} shows examples of images generated with LSGAN, WGAN
and DCGAN architectures alongside real images. The synthetic images,
particularly those produced by LSGAN, are not only highly detailed and
realistic, but also consistent with their biological nature. For example, it is
characteristic that $\beta$-Tubulin forms a circular halo cradling the nucleus.
This property is maintained clearly in most generated images.

To assess the quality of synthesized images we presented three expert biologists
with 50 real cell images and 50 artifical cells generated with LSGAN. We
randomized the order of images and asked the experts to judge whether each cell
was real or fake. On average, 30\% of the time the synthetic cells were
realistic enough to fool the human jurors into believing they were real.

>>>>

\section{Introduction}
Advances in high-throughput microscopy systems have enabled acquisition of large volumes of high-resolution cell images \cite{caicedo2017data}. This paves the way for novel computational methods that can leverage these large quantities of data to study biological systems. Our work focuses on the task of morphological profiling, which aims to map microscopy images of cells to representation vectors that capture salient features and axes of variation in an unsupervised manner \cite{caicedo_profiling}. These representations ideally divide the morphological space into clusters of cells with similar properties or, in the case of chemical or genetic treatments added to the cells, similar function.

@@

We employ Generative Adversarial Networks (GANs) \cite{goodfellow2014generative} to build a generative model of single cell images from the BBBC021 dataset \cite{Ljosa2012}. We show that this model learns rich feature representations and synthesizes realistic images that are useful for exploring morphological variation in cells. The main advantages of this method are:
\begin{itemize}
  \item \emph{Adaptability}: Because this method learns from its training data it is able to disentangle factors of variation and extract inherent semantic relationships. Transfer learning approaches lack this ability and cannot capture the intrinsic relations between the biologically meaningful channels.
  \item \emph{Translating into biological phenotypes}: The generative abilities of this model enable useful visualizations of cells to help translate data variations into meaningful biological phenotypes.
\end{itemize}

@@

\section{Exploring Biological Phenotypes Using Cell Image Synthesis}\label{synthesis}
GANs have been proven to synthesize realistic images both within and beyond the biological domain \cite{goodfellow2014generative,radford2015unsupervised,osokin2017gans}. We examine if this ability transfers to cell images extracted from the BBBC021 dataset of human breast cancer cell lines. Each image in BBBC021 consists of three \emph{channels} corresponding to DNA, F-Actin and $\beta$-Tubulin, cellular components visualized by fluorescence microscopy. We stack these channels and treat them as RGB images via a simple $\text{DNA} \mapsto \text{R}, \text{$\beta$-Tubulin} \mapsto \text{G}, \text{F-Actin} \mapsto \text{B}$ mapping. We obtain 1.3 million single cell images by segmenting the raw images using CellProfiler. Finally, we normalize the luminance of each channel to the range [0 -- 1].

Figure \ref{fig:generated} shows examples of images generated with LSGAN, WGAN
and DCGAN architectures alongside real images. The synthetic images,
particularly those produced by LSGAN, are not only highly detailed and
realistic, but also consistent with their biological nature. For example, it is
characteristic that $\beta$-Tubulin (green) forms a circular halo cradling the nucleus.
This property is maintained clearly in most generated images.

To assess the quality of synthesized images we presented three expert biologists
with 50 real cell images and 50 artificial cells generated with LSGAN. We
randomized the order of images and asked the experts to judge whether each cell
was real or fake. On average, 30\% of the time the synthetic cells were
realistic enough to fool the human jurors into believing they were real.
